\textbf{OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units}

### 2017-02-09 submission/Westcott_OptiClust_mSphere_2016.tex
Finalize documents

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

\textbf{\emph{OptiClust algorithm.}} The OptiClust algorithm uses the
pairs of sequences that are within a desired threshold of each other
(e.g.~0.03), a list of all sequence names in the dataset, and the metric
that should be used to assess clustering quality. A detailed description
of the algorithm is provided for a toy dataset in the Supplementary
Material. Briefly, the algorithm starts by placing each sequence either
within its own OTU or into a single OTU. The algorithm proceeds by
interrogating each sequence and re-calculating the metric for the cases
where the sequence stays in its current OTU, is moved to each of the
other OTUs, or is moved into a new OTU. The location that results in the
best clustering quality indicates whether the sequence should remain in
its current OTU or be moved to a different or new OTU. Each iteration
consists of interrogating every sequence in the dataset. Although
numerous options are available for optimizing the clusters and for
assessing the quality of the clusters within the mothur-based
implementation of the algorithm (e.g.~sensitivity, specificity,
accuracy, F1-score, etc.), the default metric for optimization and
assessment is MCC because it includes all four parameters from the
confusion matrix (Figure S1). The algorithm continues until the
optimization metric stabilizes or until it reaches a defined stopping
criteria.

\textbf{\emph{OptiClust-generated OTUs are more robust than those from
other methods.}} To evaluate the OptiClust algorithm and compare its
performance to other algorithms, we utilized six datasets including two
synthetic communities and four previously published large datasets
generated from soil, marine, human, and murine samples (Table 1). When
we seeded the OptiClust algorithm with each sequence in a separate OTU
and ran the algorithm until complete convergence, the MCC values
averaged 15.2 and 16.5\% higher than the OTUs using average neighbor and
distance-based greedy clustering (DGC) with VSEARCH, respectively
(Figure 1). The number of OTUs formed by the various methods was
negatively correlated with their MCC value (\(\rho\)=-0.47;
p\textless{}0.001). The OptiClust algorithm was considerably faster than
the hierarchical algorithms and somewhat slower than the heuristic-based
algorithms. Across the six datasets, the OptiClust algorithm was
94.6-times faster than average neighbor and just as fast as DGC with
VSEARCH. The human dataset was a challenge for a number of the
algorithms. OTUCLUST and SumaClust were unable to cluster the human
dataset in less than 50 hours and the average neighbor algorithm
required more than 45 GB of RAM. The USEARCH-based methods were unable
to cluster the human data using the 32-bit free version of the software
that limits the amount of RAM to approximately 3.5 GB. These data
demonstrate that OptiClust generated significantly more robust OTU
assignments than existing methods across a diverse collection of
datasets with performance that was comparable to popular methods.

@@

\textbf{\emph{Implementation of clustering algorithms.}} In addition to
the OptiClust algorithm we evaluated ten different \emph{de novo}
clustering algorithms. These included three hierarchical algorithms,
average neighbor, nearest neighbor, and furthest neighbor, which are
implemented in mothur (v.1.39.0) (11). Seven heuristic methods were also
used including abundance-based greedy clustering (AGC) and
(distance-based greedy clustering) DGC as implemented in USEARCH (v.6.1)
(17) and VSEARCH (v.2.3.3) ((18){]}, OTUCLUST (v.0.1) (19), SumaClust
(v.1.0.20), and Swarm (v.2.1.9) (20). With the exception of Swarm each
of these methods uses distance-based thresholds to report OTU
assignments. We also evalauted the ability of OptiClust to optimize to
metrics other than MCC. These included accuracy, F1-score, negative
predictive value, positive predictive value, false discovery rate,
senitivity, specificity, the sum of TPs and TNs, the sum of FPs and FNs,
and the number of FNs, FPs, TNs, and TPs (Figure S1).

@@

\textbf{\emph{Data and code availability.}} The workflow utilized
commands in GNU make (v.3.81), GNU bash (v.4.1.2), mothur (v.1.39.0)
(11), and R (v.3.3.2) (44). Within R we utilized the wesanderson
(v.0.3.2) (45), dplyr (v.0.5.0) (46), tidyr (v.0.6.0) (47), cowplot
(v.0.6.9990) (48), and ggplot2 (v.2.1.0.9001) (49) packages. A
reproducible version of this manuscript and analysis is available at
\url{https://github.com/SchlossLab/Westcott_OptiClust_mSphere_2017}.

@@

@@

\hyperdef{}{references}{\label{references}}
\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hyperdef{}{ref-Schloss2016b}{\label{ref-Schloss2016b}}
1. \textbf{Schloss PD}, \textbf{Girard RA}, \textbf{Martin T},
\textbf{Edwards J}, \textbf{Thrash JC}. 2016. Status of the archaeal and
bacterial census: An update. mBio \textbf{7}:e00201--16.
doi:\url{http://doi.org/10.1128/mbio.00201-16}.

\hyperdef{}{ref-Locey2016}{\label{ref-Locey2016}}
2. \textbf{Locey KJ}, \textbf{Lennon JT}. 2016. Scaling laws predict
global microbial diversity. Proceedings of the National Academy of
Sciences \textbf{113}:5970--5975.
doi:\url{http://doi.org/10.1073/pnas.1521291113}.

\hyperdef{}{ref-Rideout2014}{\label{ref-Rideout2014}}
3. \textbf{Rideout JR}, \textbf{He Y}, \textbf{Navas-Molina JA},
\textbf{Walters WA}, \textbf{Ursell LK}, \textbf{Gibbons SM},
\textbf{Chase J}, \textbf{McDonald D}, \textbf{Gonzalez A},
\textbf{Robbins-Pianka A}, \textbf{Clemente JC}, \textbf{Gilbert JA},
\textbf{Huse SM}, \textbf{Zhou H-W}, \textbf{Knight R}, \textbf{Caporaso
JG}. 2014. Subsampled open-reference clustering creates consistent,
comprehensive OTU definitions and scales to billions of sequences. PeerJ
\textbf{2}:e545. doi:\url{http://doi.org/10.7717/peerj.545}.

\hyperdef{}{ref-Huttenhower2012}{\label{ref-Huttenhower2012}}
4. \textbf{Consortium THMP}. 2012. Structure, function and diversity of
the healthy human microbiome. Nature \textbf{486}:207--214.
doi:\url{http://doi.org/10.1038/nature11234}.

\hyperdef{}{ref-Eckburg2005}{\label{ref-Eckburg2005}}
5. \textbf{Eckburg PB}, \textbf{Bik EM}, \textbf{Bernstein CN},
\textbf{Purdom E}, \textbf{Dethlefsen L}, \textbf{Sargent M},
\textbf{Gill SR}, \textbf{Nelson KE}, \textbf{Relman DA}. 2005.
Diversity of the human intestinal microbial flora. Science
\textbf{308}:1635--1638.

\hyperdef{}{ref-Elshahed2008}{\label{ref-Elshahed2008}}
6. \textbf{Elshahed MS}, \textbf{Youssef NH}, \textbf{Spain AM},
\textbf{Sheik C}, \textbf{Najar FZ}, \textbf{Sukharnikov LO},
\textbf{Roe BA}, \textbf{Davis JP}, \textbf{Schloss PD}, \textbf{Bailey
VL}, \textbf{Krumholz LR}. 2008. Novelty and uniqueness patterns of rare
members of the soil biosphere. Applied and Environmental Microbiology
\textbf{74}:5422--5428. doi:\url{http://doi.org/10.1128/aem.00410-08}.

\hyperdef{}{ref-Kozich2013}{\label{ref-Kozich2013}}
7. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq Illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\url{http://doi.org/10.1128/aem.01043-13}.

\hyperdef{}{ref-Schloss2009a}{\label{ref-Schloss2009a}}
8. \textbf{Schloss PD}. 2009. A high-throughput DNA sequence aligner for
microbial ecology studies. PLOS ONE \textbf{4}:e8230.
doi:\url{http://doi.org/10.1371/journal.pone.0008230}.

\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
9. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
doi:\url{http://doi.org/10.1093/bioinformatics/btr381}.

\hyperdef{}{ref-Wang2007}{\label{ref-Wang2007}}
10. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
of rRNA sequences into the new bacterial taxonomy. Applied and
Environmental Microbiology \textbf{73}:5261--5267.
doi:\url{http://doi.org/10.1128/aem.00062-07}.

\hyperdef{}{ref-Schloss2009b}{\label{ref-Schloss2009b}}
11. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541. doi:\url{http://doi.org/10.1128/aem.01541-09}.

\hyperdef{}{ref-Caporaso2010}{\label{ref-Caporaso2010}}
12. \textbf{Caporaso JG}, \textbf{Kuczynski J}, \textbf{Stombaugh J},
\textbf{Bittinger K}, \textbf{Bushman FD}, \textbf{Costello EK},
\textbf{Fierer N}, \textbf{Peña AG}, \textbf{Goodrich JK},
\textbf{Gordon JI}, \textbf{Huttley GA}, \textbf{Kelley ST},
\textbf{Knights D}, \textbf{Koenig JE}, \textbf{Ley RE},
\textbf{Lozupone CA}, \textbf{McDonald D}, \textbf{Muegge BD},
\textbf{Pirrung M}, \textbf{Reeder J}, \textbf{Sevinsky JR},
\textbf{Turnbaugh PJ}, \textbf{Walters WA}, \textbf{Widmann J},
\textbf{Yatsunenko T}, \textbf{Zaneveld J}, \textbf{Knight R}. 2010.
QIIME allows analysis of high-throughput community sequencing data.
Nature Methods \textbf{7}:335--336.
doi:\url{http://doi.org/10.1038/nmeth.f.303}.

\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
13. \textbf{Schloss PD}, \textbf{Westcott SL}. 2011. Assessing and
improving methods used in operational taxonomic unit-based approaches
for 16S rRNA gene sequence analysis. Applied and Environmental
Microbiology \textbf{77}:3219--3226.
doi:\url{http://doi.org/10.1128/aem.02810-10}.

\hyperdef{}{ref-NavasMolina2013}{\label{ref-NavasMolina2013}}
14. \textbf{Navas-Molina JA}, \textbf{Peralta-Sánchez JM},
\textbf{González A}, \textbf{McMurdie PJ}, \textbf{Vázquez-Baeza Y},
\textbf{Xu Z}, \textbf{Ursell LK}, \textbf{Lauber C}, \textbf{Zhou H},
\textbf{Song SJ}, \textbf{Huntley J}, \textbf{Ackermann GL},
\textbf{Berg-Lyons D}, \textbf{Holmes S}, \textbf{Caporaso JG},
\textbf{Knight R}. 2013. Advancing our understanding of the human
microbiome using QIIME, pp. 371--444. \emph{In} Methods in enzymology.
Elsevier BV.

\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
15. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
methods outperform reference-based methods for assigning 16S rRNA gene
sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
doi:\url{http://doi.org/10.7717/peerj.1487}.

\hyperdef{}{ref-Schloss2005}{\label{ref-Schloss2005}}
16. \textbf{Schloss PD}, \textbf{Handelsman J}. 2005. Introducing DOTUR,
a computer program for defining operational taxonomic units and
estimating species richness. Applied and Environmental microbiology
\textbf{71}:1501--1506.

\hyperdef{}{ref-Edgar2010}{\label{ref-Edgar2010}}
17. \textbf{Edgar RC}. 2010. Search and clustering orders of magnitude
faster than BLAST. Bioinformatics \textbf{26}:2460--2461.
doi:\url{http://doi.org/10.1093/bioinformatics/btq461}.

\hyperdef{}{ref-Rognes2016}{\label{ref-Rognes2016}}
18. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\url{http://doi.org/10.7717/peerj.2584}.

\hyperdef{}{ref-Albanese2015}{\label{ref-Albanese2015}}
19. \textbf{Albanese D}, \textbf{Fontana P}, \textbf{Filippo CD},
\textbf{Cavalieri D}, \textbf{Donati C}. 2015. MICCA: A complete and
accurate software for taxonomic profiling of metagenomic data.
Scientific Reports \textbf{5}:9743.
doi:\url{http://doi.org/10.1038/srep09743}.

\hyperdef{}{ref-Mah2014}{\label{ref-Mah2014}}
20. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
\textbf{Vargas C de}, \textbf{Dunthorn M}. 2014. Swarm: Robust and fast
clustering method for amplicon-based studies. PeerJ \textbf{2}:e593.
doi:\url{http://doi.org/10.7717/peerj.593}.

\hyperdef{}{ref-Sun2009}{\label{ref-Sun2009}}
21. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Liu L}, \textbf{Yu F},
\textbf{Farrell ML}, \textbf{McKendree W}, \textbf{Farmerie W}. 2009.
ESPRIT: Estimating species richness using large collections of 16S rRNA
pyrosequences. Nucleic Acids Research \textbf{37}:e76--e76.
doi:\url{http://doi.org/10.1093/nar/gkp285}.

\hyperdef{}{ref-Cai2011}{\label{ref-Cai2011}}
22. \textbf{Cai Y}, \textbf{Sun Y}. 2011. ESPRIT-tree: Hierarchical
clustering analysis of millions of 16S rRNA pyrosequences in quasilinear
computational time. Nucleic Acids Research \textbf{39}:e95--e95.
doi:\url{http://doi.org/10.1093/nar/gkr349}.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

\textbf{\emph{OptiClust algorithm.}} The OptiClust algorithm uses the
pairs of sequences that are within a desired threshold of each other
(e.g.~0.03), a list of all sequence names in the dataset, and the metric
that should be used to assess clustering quality. A detailed description
of the algorithm is provided for a toy dataset in the Supplementary
Material. Briefly, the algorithm starts by placing each sequence either
within its own OTU or into a single OTU. The algorithm proceeds by
interrogating each sequence and re-calculating the metric for the cases
where the sequence stays in its current OTU, is moved to each of the
other OTUs, or is moved into a new OTU. The location that results in the
best clustering quality indicates whether the sequence should remain in
its current OTU or be moved to a different or new OTU. Each iteration
consists of interrogating every sequence in the dataset. Although
numerous options are available for optimizing the clusters and for
assessing the quality of the clusters within the mothur-based
implementation of the algorithm (e.g.~sensitivity, specificity,
accuracy, F1-score, etc.), the default metric for optimization and
assessment is MCC because it includes all four parameters from the
confusion matrix (Figure S1; Table S1). The algorithm continues until
the optimization metric stabilizes or until it reaches a defined
stopping criteria.

\textbf{\emph{OptiClust-generated OTUs are more robust than those from
other methods.}} To evaluate the OptiClust algorithm and compare its
performance to other algorithms, we utilized six datasets including two
synthetic communities and four previously published large datasets
generated from soil, marine, human, and murine samples (Table 1). When
we seeded the OptiClust algorithm with each sequence in a separate OTU
and ran the algorithm until complete convergence, the MCC values
averaged 15.2 and 16.5\% higher than the OTUs using average neighbor and
distance-based greedy clustering (DGC) with VSEARCH, respectively
(Figure 1; Table S1). The number of OTUs formed by the various methods
was negatively correlated with their MCC value (\(\rho\)=-0.47;
p\textless{}0.001). The OptiClust algorithm was considerably faster than
the hierarchical algorithms and somewhat slower than the heuristic-based
algorithms. Across the six datasets, the OptiClust algorithm was
94.6-times faster than average neighbor and just as fast as DGC with
VSEARCH. The human dataset was a challenge for a number of the
algorithms. OTUCLUST and SumaClust were unable to cluster the human
dataset in less than 50 hours and the average neighbor algorithm
required more than 45 GB of RAM. The USEARCH-based methods were unable
to cluster the human data using the 32-bit free version of the software
that limits the amount of RAM to approximately 3.5 GB. These data
demonstrate that OptiClust generated significantly more robust OTU
assignments than existing methods across a diverse collection of
datasets with performance that was comparable to popular methods.

@@

\textbf{\emph{Implementation of clustering algorithms.}} In addition to
the OptiClust algorithm we evaluated ten different \emph{de novo}
clustering algorithms. These included three hierarchical algorithms,
average neighbor, nearest neighbor, and furthest neighbor, which are
implemented in mothur (v.1.39.0) (11). Seven heuristic methods were also
used including abundance-based greedy clustering (AGC) and
(distance-based greedy clustering) DGC as implemented in USEARCH (v.6.1)
(17) and VSEARCH (v.2.3.3) ((18){]}, OTUCLUST (v.0.1) (19), SumaClust
(v.1.0.20), and Swarm (v.2.1.9) (20). With the exception of Swarm each
of these methods uses distance-based thresholds to report OTU
assignments. We also evalauted the ability of OptiClust to optimize to
metrics other than MCC. These included accuracy, F1-score, negative
predictive value, positive predictive value, false discovery rate,
senitivity, specificity, the sum of TPs and TNs, the sum of FPs and FNs,
and the number of FNs, FPs, TNs, and TPs (Figure S1; Table S1).

@@

\textbf{\emph{Data and code availability.}} The workflow utilized
commands in GNU make (v.3.81), GNU bash (v.4.1.2), mothur (v.1.39.0)
(11), and R (v.3.3.2) (44). Within R we utilized the wesanderson
(v.0.3.2) (45), dplyr (v.0.5.0) (46), tidyr (v.0.6.0) (47), cowplot
(v.0.6.3) (48), and ggplot2 (v.2.2.0.9000) (49) packages. A reproducible
version of this manuscript and analysis is available at
\url{https://github.com/SchlossLab/Westcott_OptiClust_mSphere_2017}.

@@

\textbf{Table S1. Summary of the average number of true positives, true
negatives, false positives, false negatives and the resulting Matthews
correlation coefficient for each of the clustering methods that were
analyzed in this study for each of the six datasets.} Blank values
indicate that those conditions could not be completed in 50 hours with
45 GB of RAM.

@@

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-Schloss2016b}{}
1. \textbf{Schloss PD}, \textbf{Girard RA}, \textbf{Martin T},
\textbf{Edwards J}, \textbf{Thrash JC}. 2016. Status of the archaeal and
bacterial census: An update. mBio \textbf{7}:e00201--16.
doi:\href{https://doi.org/10.1128/mbio.00201-16}{10.1128/mbio.00201-16}.

\hypertarget{ref-Locey2016}{}
2. \textbf{Locey KJ}, \textbf{Lennon JT}. 2016. Scaling laws predict
global microbial diversity. Proceedings of the National Academy of
Sciences \textbf{113}:5970--5975.
doi:\href{https://doi.org/10.1073/pnas.1521291113}{10.1073/pnas.1521291113}.

\hypertarget{ref-Rideout2014}{}
3. \textbf{Rideout JR}, \textbf{He Y}, \textbf{Navas-Molina JA},
\textbf{Walters WA}, \textbf{Ursell LK}, \textbf{Gibbons SM},
\textbf{Chase J}, \textbf{McDonald D}, \textbf{Gonzalez A},
\textbf{Robbins-Pianka A}, \textbf{Clemente JC}, \textbf{Gilbert JA},
\textbf{Huse SM}, \textbf{Zhou H-W}, \textbf{Knight R}, \textbf{Caporaso
JG}. 2014. Subsampled open-reference clustering creates consistent,
comprehensive OTU definitions and scales to billions of sequences. PeerJ
\textbf{2}:e545.
doi:\href{https://doi.org/10.7717/peerj.545}{10.7717/peerj.545}.

\hypertarget{ref-Huttenhower2012}{}
4. \textbf{Consortium THMP}. 2012. Structure, function and diversity of
the healthy human microbiome. Nature \textbf{486}:207--214.
doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.

\hypertarget{ref-Eckburg2005}{}
5. \textbf{Eckburg PB}, \textbf{Bik EM}, \textbf{Bernstein CN},
\textbf{Purdom E}, \textbf{Dethlefsen L}, \textbf{Sargent M},
\textbf{Gill SR}, \textbf{Nelson KE}, \textbf{Relman DA}. 2005.
Diversity of the human intestinal microbial flora. Science
\textbf{308}:1635--1638.

\hypertarget{ref-Elshahed2008}{}
6. \textbf{Elshahed MS}, \textbf{Youssef NH}, \textbf{Spain AM},
\textbf{Sheik C}, \textbf{Najar FZ}, \textbf{Sukharnikov LO},
\textbf{Roe BA}, \textbf{Davis JP}, \textbf{Schloss PD}, \textbf{Bailey
VL}, \textbf{Krumholz LR}. 2008. Novelty and uniqueness patterns of rare
members of the soil biosphere. Applied and Environmental Microbiology
\textbf{74}:5422--5428.
doi:\href{https://doi.org/10.1128/aem.00410-08}{10.1128/aem.00410-08}.

\hypertarget{ref-Kozich2013}{}
7. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq Illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-Schloss2009a}{}
8. \textbf{Schloss PD}. 2009. A high-throughput DNA sequence aligner for
microbial ecology studies. PLOS ONE \textbf{4}:e8230.
doi:\href{https://doi.org/10.1371/journal.pone.0008230}{10.1371/journal.pone.0008230}.

\hypertarget{ref-Edgar2011}{}
9. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.

\hypertarget{ref-Wang2007}{}
10. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
of rRNA sequences into the new bacterial taxonomy. Applied and
Environmental Microbiology \textbf{73}:5261--5267.
doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.

\hypertarget{ref-Schloss2009b}{}
11. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Caporaso2010}{}
12. \textbf{Caporaso JG}, \textbf{Kuczynski J}, \textbf{Stombaugh J},
\textbf{Bittinger K}, \textbf{Bushman FD}, \textbf{Costello EK},
\textbf{Fierer N}, \textbf{Peña AG}, \textbf{Goodrich JK},
\textbf{Gordon JI}, \textbf{Huttley GA}, \textbf{Kelley ST},
\textbf{Knights D}, \textbf{Koenig JE}, \textbf{Ley RE},
\textbf{Lozupone CA}, \textbf{McDonald D}, \textbf{Muegge BD},
\textbf{Pirrung M}, \textbf{Reeder J}, \textbf{Sevinsky JR},
\textbf{Turnbaugh PJ}, \textbf{Walters WA}, \textbf{Widmann J},
\textbf{Yatsunenko T}, \textbf{Zaneveld J}, \textbf{Knight R}. 2010.
QIIME allows analysis of high-throughput community sequencing data.
Nature Methods \textbf{7}:335--336.
doi:\href{https://doi.org/10.1038/nmeth.f.303}{10.1038/nmeth.f.303}.

\hypertarget{ref-Schloss2011}{}
13. \textbf{Schloss PD}, \textbf{Westcott SL}. 2011. Assessing and
improving methods used in operational taxonomic unit-based approaches
for 16S rRNA gene sequence analysis. Applied and Environmental
Microbiology \textbf{77}:3219--3226.
doi:\href{https://doi.org/10.1128/aem.02810-10}{10.1128/aem.02810-10}.

\hypertarget{ref-NavasMolina2013}{}
14. \textbf{Navas-Molina JA}, \textbf{Peralta-Sánchez JM},
\textbf{González A}, \textbf{McMurdie PJ}, \textbf{Vázquez-Baeza Y},
\textbf{Xu Z}, \textbf{Ursell LK}, \textbf{Lauber C}, \textbf{Zhou H},
\textbf{Song SJ}, \textbf{Huntley J}, \textbf{Ackermann GL},
\textbf{Berg-Lyons D}, \textbf{Holmes S}, \textbf{Caporaso JG},
\textbf{Knight R}. 2013. Advancing our understanding of the human
microbiome using QIIME, pp. 371--444. \emph{In} Methods in enzymology.
Elsevier BV.

\hypertarget{ref-Westcott2015}{}
15. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
methods outperform reference-based methods for assigning 16S rRNA gene
sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.

\hypertarget{ref-Schloss2005}{}
16. \textbf{Schloss PD}, \textbf{Handelsman J}. 2005. Introducing DOTUR,
a computer program for defining operational taxonomic units and
estimating species richness. Applied and Environmental microbiology
\textbf{71}:1501--1506.

\hypertarget{ref-Edgar2010}{}
17. \textbf{Edgar RC}. 2010. Search and clustering orders of magnitude
faster than BLAST. Bioinformatics \textbf{26}:2460--2461.
doi:\href{https://doi.org/10.1093/bioinformatics/btq461}{10.1093/bioinformatics/btq461}.

\hypertarget{ref-Rognes2016}{}
18. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.

\hypertarget{ref-Albanese2015}{}
19. \textbf{Albanese D}, \textbf{Fontana P}, \textbf{Filippo CD},
\textbf{Cavalieri D}, \textbf{Donati C}. 2015. MICCA: A complete and
accurate software for taxonomic profiling of metagenomic data.
Scientific Reports \textbf{5}:9743.
doi:\href{https://doi.org/10.1038/srep09743}{10.1038/srep09743}.

\hypertarget{ref-Mah2014}{}
20. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
\textbf{Vargas C de}, \textbf{Dunthorn M}. 2014. Swarm: Robust and fast
clustering method for amplicon-based studies. PeerJ \textbf{2}:e593.
doi:\href{https://doi.org/10.7717/peerj.593}{10.7717/peerj.593}.

\hypertarget{ref-Sun2009}{}
21. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Liu L}, \textbf{Yu F},
\textbf{Farrell ML}, \textbf{McKendree W}, \textbf{Farmerie W}. 2009.
ESPRIT: Estimating species richness using large collections of 16S rRNA
pyrosequences. Nucleic Acids Research \textbf{37}:e76--e76.
doi:\href{https://doi.org/10.1093/nar/gkp285}{10.1093/nar/gkp285}.

\hypertarget{ref-Cai2011}{}
22. \textbf{Cai Y}, \textbf{Sun Y}. 2011. ESPRIT-tree: Hierarchical
clustering analysis of millions of 16S rRNA pyrosequences in quasilinear
computational time. Nucleic Acids Research \textbf{39}:e95--e95.
doi:\href{https://doi.org/10.1093/nar/gkr349}{10.1093/nar/gkr349}.

\hypertarget{ref-Edgar2013}{}
23. \textbf{Edgar RC}. 2013. UPARSE: Highly accurate OTU sequences from
microbial amplicon reads. Nature Methods \textbf{10}:996--998.
doi:\href{https://doi.org/10.1038/nmeth.2604}{10.1038/nmeth.2604}.

\hypertarget{ref-Mah2015}{}
24. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
\textbf{Vargas C de}, \textbf{Dunthorn M}. 2015. Swarm v2:
Highly-scalable and high-resolution amplicon clustering. PeerJ
\textbf{3}:e1420.
doi:\href{https://doi.org/10.7717/peerj.1420}{10.7717/peerj.1420}.
